\section{Introduction}

Generative models increasingly contribute text, images, code, and other artifacts to the same information environment from which later models may be trained. This creates a model--data feedback loop: a model trained on a human-origin distribution generates synthetic examples; a successor is trained on a dataset containing those examples; and the process repeats. Several controlled studies show that particular versions of this loop can progressively distort a learned distribution, increase held-out error, reduce diversity, or lose low-probability events \cite{shumailov2024collapse,alemohammad2024mad,briesch2023llmoutput,dohmatob2024tails,dohmatob2024regression,suresh2025rate}. We use the cautious term \emph{recursive-training degradation} because ``model collapse'' has been defined through different endpoints, including tail loss, variance contraction, divergent test error, and declining task performance.

The literature also shows that degradation is not an unconditional consequence of using synthetic data. Outcomes depend on how data are generated, filtered, retained, and mixed. Accumulating original and synthetic data can keep error bounded in settings where replacing the dataset causes degradation \cite{gerstgrasser2024inevitable,kazdan2025collapse}. Fixed real-data fractions, token editing, external correction, verification, and detector-based importance resampling can also stabilize some recursive procedures \cite{seddik2024synthetic,gillman2024selfcorrecting,drayson2025machine,zhu2025synthesize,zhu2026reflow,yi2026verification}. These findings shift the useful question away from whether synthetic data is universally safe or unsafe. The practical question is which data-use policy is appropriate under an explicit resource constraint.

Human-origin data can be costly to license, curate, deduplicate, annotate, or reserve from final evaluation. Even when a large archive exists, repeated optimizer presentation of that archive is not free: it consumes training tokens and can create an unfair comparison against a synthetic-only policy. We therefore treat human-origin training data as a finite lifetime resource. The budget is measured in \emph{optimizer-consumed tokens}, so presenting a human example three times consumes three times its token length. This definition prevents a policy from appearing efficient merely because repeated epochs over the same human records are counted as one use.

A lifetime budget introduces a scheduling problem. Spending human tokens early may slow the propagation of an error before it compounds, whereas spending them late may target a state that has already revealed where degradation is concentrated. The budget also introduces a selection problem. Randomly sampled human text may reinforce already well-covered regions while rare or under-covered modes continue to drift. Conversely, targeting the most surprising or weakly represented mode can overfit a biased monitor, chase noise, or neglect an unobserved but globally important mode. Prior work separately motivates data accumulation and fresh mixing, learned domain mixtures, dynamic schedules, importance resampling, active selection, and tail-aware robustness \cite{xie2023doremi,ye2025mixinglaws,liu2025regmix,zhao2026regmixd,wang2026tikmix,xie2023dsir,ash2020badge,liu2021jtt,sohoni2020george}. Recent dynamic-mixture methods are especially important novelty threats because they vary both training stage and data domain, although they do not study a recursive human--synthetic chain with one fixed human-origin lifetime stock.

This paper asks the following result-free research question: \emph{under a fixed lifetime budget of human-origin optimizer tokens and a fixed total optimizer-token budget, does jointly allocating human tokens across recursive generations and monitored human-distribution modes reduce chain-level degradation relative to the strongest eligible schedule-only and selection-only policies?} The question is intentionally comparative. A joint policy is not credited for beating only a no-rescue chain or a weak random baseline. It must be compared with policies that can optimize when to spend while sampling modes neutrally, and policies that can target modes while following a fixed schedule.

Let the recursive horizon be $G$, let a frozen monitoring map partition the human reference into $M$ modes, and let $B$ be the lifetime human-origin optimizer-token budget. A policy chooses allocations $b_{g,m}$ with $\sum_{g=1}^{G}\sum_{m=1}^{M} b_{g,m}=B$. The primary experimental unit is a complete independently seeded recursive chain, not an individual generation. Proposed primary outcomes aggregate held-out human negative-log-likelihood regret and a separately frozen tail-retention measure across the full horizon. Final human test data remain unavailable to the policy, model selection, threshold setting, and early stopping.

A fixed lifetime budget also changes what constitutes a fair baseline. A policy that receives the same human fraction at every generation may consume far more human-origin tokens than a policy operating for fewer generations, while a policy that replays a small archive can quietly obtain many optimizer exposures. Our comparisons therefore match both the lifetime human-token total and the overall optimizer-token total. This choice does not model every real acquisition cost, but it isolates whether allocation itself matters once training exposure is controlled. It also makes a negative result informative: if a simple fixed schedule or untargeted mixture matches the joint policy under the same accounting, the added adaptivity is not justified in the tested regime.

The planned contribution is limited to four items. First, we formalize the joint time-and-mode allocation problem under exact lifetime human-token and total-token matching. Second, we define a decomposition into random, schedule-only, selection-only, and joint treatment families. Third, we predeclare a monitoring-omission intervention that can reveal when targeted rescue optimizes the wrong reference. Fourth, we provide a chain-level evaluation protocol that permits a null or harmful outcome. No experiment in this repository currently establishes that joint allocation helps, prevents collapse, preserves rare modes, or is optimal. The novelty statement is provisional and must be withdrawn if a closer primary source is found.
