\section{Related Work}

\paragraph{Recursive degradation and collapse.}
Early controlled studies established that repeated fitting to model-generated samples can amplify approximation and sampling errors. Shumailov et al. distinguish early tail loss from late distributional collapse across simple distributions, image models, and language models \cite{shumailov2024collapse}. Alemohammad et al. analyze self-consuming image-generation loops with fixed or fresh real data and show quality--diversity deterioration when fresh support is insufficient \cite{alemohammad2024mad}. Briesch et al. report declining output diversity in recursive language-model loops, with sensitivity to the synthetic-data share and initial diversity \cite{briesch2023llmoutput}. Theoretical work makes the dependence on estimator, sample size, spectrum, and horizon explicit. Dohmatob et al. derive recursive regression error and scaling-law changes \cite{dohmatob2024regression,dohmatob2024tails}, while Suresh et al. characterize the time required to lose discrete symbols and contract Gaussian variance \cite{suresh2025rate}. These papers motivate tracking the complete trajectory and tail behavior rather than reporting only an endpoint.

The recursive setting is heterogeneous. Human-source properties can change both the magnitude and direction of distribution shift, including lexical or semantic diversity and political bias \cite{kovac2025recursive}. Self-consuming performative loops can also alter group bias through the interaction of model outputs and subsequent data collection \cite{wang2026scpl}. Accordingly, this project does not treat a single scalar ``collapse'' score as universal. It freezes one domain, one recursive protocol, and explicit held-out human outcomes.

\paragraph{Fresh mixing, accumulation, and synthetic-data construction.}
Several works show that retaining or injecting non-synthetic information changes the dynamics. Gerstgrasser et al. compare replacement with accumulation and find bounded or stable error under accumulation in language, molecular, image, and linear settings \cite{gerstgrasser2024inevitable}. Kazdan et al. extend the comparison to replacement, full accumulation, and fixed-size subsets sampled from accumulated pools, showing that the data-use workflow can determine whether degradation is explosive, gradual, or contained \cite{kazdan2025collapse}. Seddik et al. derive conditions on the real--synthetic mixture under which collapse can be avoided in a statistical language-model setting \cite{seddik2024synthetic}. Kang et al. study natural, rephrased, and textbook-style synthetic text over model and data scales, showing that useful mixture ratios depend on synthetic-data type, model size, and budget \cite{kang2025demystifying}.

Other interventions alter synthetic examples or add an external information channel. Gillman et al. use idealized and domain-specific correction functions in self-consuming loops \cite{gillman2024selfcorrecting}. Zhu et al. propose token editing of human-produced text to construct semi-synthetic data with bounded-error motivation \cite{zhu2025synthesize}. In Reflow-style recursion, a fixed fraction of real data prevents the contraction identified by a linear denoising-autoencoder analysis and improves practical stability \cite{zhu2026reflow}. Yi et al. study verifier-guided retraining and emphasize that an imperfect verifier can determine the long-run knowledge center \cite{yi2026verification}. These methods demonstrate that external information can be valuable, but they do not answer how to allocate one exhaustible human-token stock across both generation and mode.

\paragraph{Scheduling and data-mixture optimization.}
Data-mixture research treats domains as controllable components of pretraining. DoReMi learns static domain weights with a proxy model and group distributionally robust optimization \cite{xie2023doremi}. Data Mixing Laws predict performance as a function of mixture proportions and extend the analysis to continual training \cite{ye2025mixinglaws}. RegMix fits a regression model to many small proxy runs and searches for a high-performing static mixture \cite{liu2025regmix}. These methods provide strong mode-allocation analogues even though their domains are not necessarily human-distribution modes and their training is not recursive.

Recent work makes the mixture time-dependent. RegMix-D uses proxy loss trajectories to construct offline or online schedules that switch domain proportions during target training \cite{zhao2026regmixd}. TiKMiX periodically updates domain weights using group influence while limiting the overhead of fully online adaptation \cite{wang2026tikmix}. These are high-threat closest works: they establish that jointly varying training time and domain is not new in general. The remaining provisional distinction of our problem is the conjunction of a recursive human--synthetic chain, exact human-origin optimizer-token accounting, one fixed lifetime human stock, frozen monitored modes, and matched schedule-only and selection-only controls.

The scheduling literature therefore supplies more than background terminology. It supplies concrete competitors and a warning about overclaiming: changing data proportions over training time is already an established design pattern. For this project, a schedule-only policy should be allowed to use the strongest state signal available without changing within-generation mode proportions, and a selection-only policy should receive the strongest frozen budget schedule. Only the residual contrast against those controls can speak to the value of coupling the two decisions.

\paragraph{Selection, surprise, and semantic coverage.}
Importance and active sampling provide candidate mechanisms for deciding which human examples to spend. DSIR estimates density ratios in a reduced feature space and resamples a large raw corpus to match an unlabeled target distribution \cite{xie2023dsir}. Katharopoulos and Fleuret focus optimization on examples with high estimated gradient contribution while preserving an unbiased objective \cite{katharopoulos2018importance}. Active Bias weights samples using prediction-history variance or threshold proximity \cite{chang2017activebias}. Core-set active learning selects geometrically representative points \cite{sener2018coreset}, whereas BADGE combines uncertainty and diversity in gradient-embedding space \cite{ash2020badge}. In recursive training, Drayson et al. train a human-versus-machine detector and importance-resample likely human text, making it a particularly relevant selection-only baseline \cite{drayson2025machine}. None of these selectors is assumed to be novel here. A proposed mode score must be frozen, compared with strong alternatives, and allowed to fail.

\paragraph{Tail preservation and monitoring bias.}
Average loss can conceal degradation on rare or latent groups. Group DRO optimizes worst-group loss when group labels are known, with regularization required for worst-group generalization in overparameterized networks \cite{sagawa2020groupdro}. JTT identifies high-loss examples with an initial model and upsamples them in a second training stage \cite{liu2021jtt}. GEORGE clusters learned representations to estimate hidden subclasses and then applies robust training, directly illustrating that monitoring labels can be missing or misspecified \cite{sohoni2020george}. Logit adjustment corrects long-tailed label-frequency bias when the relevant labels and frequencies are known \cite{menon2021logit}. These works motivate a tail-retention outcome, but also warn against equating high loss with the desired minority mode.

Our targeted policy observes only a frozen monitoring reference. If that reference omits a globally important mode, a policy can appear locally efficient while worsening true held-out performance. The planned monitoring-omission intervention therefore serves as a falsification test, not a robustness decoration. It separates the question ``can a selector exploit a good monitor?'' from the stronger and potentially false claim ``does targeting reliably preserve the human distribution?''
