\section{Limitations}

This study is designed as a focused pilot rather than a general solution to recursive training. The planned primary evidence covers one licensed text domain and a small screening-scale language model. Behavior at this scale may not transfer to frontier pretraining, other model families, multimodal systems, instruction tuning, or production data pipelines. Even a consistent result across the planned chains would support only the frozen domain, model, horizon, decoding rule, and budget regime.

The monitored modes are constructed rather than naturally complete. Semantic clusters or available labels may merge distinct subpopulations, split a coherent mode, or omit a socially or scientifically important tail. A targeted policy can therefore be efficient relative to its monitor while being harmful relative to the full human distribution. The monitoring-omission intervention probes one form of this failure, but it cannot enumerate all misspecification. Mode-specific statistics are also noisy at small sample sizes, and an adaptive policy may spend budget in response to estimation error rather than genuine degradation.

Exact budget matching creates its own simplifications. Counting optimizer-consumed tokens is more auditable than counting documents, but equal token counts do not guarantee equal informational value, annotation cost, licensing cost, duplication, or compute. Repeated human examples may be less valuable than fresh examples even when their token presentations are counted equally. Conversely, synthetic examples can vary widely in quality and generation cost. The pilot does not establish an economically optimal human-data policy.

Compute limits constrain both the number of independent chains and the breadth of baselines. Generations within a chain are dependent observations and cannot substitute for additional seeds. If the affordable seed count yields wide uncertainty, the correct conclusion may be inconclusive. Screening a schedule or selector on the same small system can also overfit policy hyperparameters. The project must retain failed runs, apply objective invalidation rules, and avoid replacing divergent chains merely because they weaken the hypothesis.

The literature contains strong alternatives: accumulation, fixed real-data mixing, correction or verification, detector-based resampling, and dynamic domain-mixture optimization. The proposed joint policy may offer no advantage over these simpler methods, and the pilot may be unable to implement every baseline at publication quality. Any omitted strong baseline narrows the novelty and performance claims. In particular, this work does not claim that synthetic data inevitably causes collapse, that human data is always superior, that time-and-mode allocation is new in general, or that the proposed policy is optimal.

Finally, this manuscript is result-free. No validated novel recursive-chain result exists in the repository at the time of writing. The method, exact domain, monitoring map, effect threshold, and powered seed count remain subject to preregistration gates. Conclusions must report null, harmful, or ambiguous outcomes if those are what the immutable artifacts support.
